---
title: "AI/DS Repos"
author: "Augustina Ragwitz"
date: "June 21, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r includes, message=FALSE}
library(lubridate)
library(tidyverse)

```

# Question

Concerning open source engagement, are companies engaging more in their own projects than in external ones? Are companies collaborating or working in isolate?

This analysis specifically focuses on Artificial Intelligence and Data Science projects.

# Data

The data come from other analysis, detailed below.

## Projects of Interest

From countering-bean-counting/git-commit-log-engagement/deep-learning-frameworks-commit-log/dl-frameworks-commit-log.Rmd

```{r gitlog-open-projects}

# manual list of sponsored projects
sponsored_projects <- read_csv('
tensorflow,Google
paddlepaddle,Baidu
pytorch,Facebook
keras-team,Google
', col_names=c("project","sponsor"))

# Github Repositories
repos_open <- read_rds("imported_data/repos_open.Rds")

repos_open <- repos_open %>% 
  select(project=owner.login, repo=full_name, description) %>%
  mutate(commit_access = "open", project=str_to_lower(project)) %>%
  left_join(sponsored_projects)

# Commit Log for Repositories (last 6 months)
gitlog_open_projects <- read_rds("imported_data/commit_log_open_Jun2018.Rds")

gitlog_open_projects <- gitlog_open_projects %>%
  select(project=org,
         repo,
         commit_month, 
         author_name, author_email,
         committer_name, committer_email,
         sha,
         committer_company_name, committer_company_type,
         author_company_name, author_company_type)

```

## Companies of Interest

```{r gitlog-closed-projects}

# Github Repositories
repos_closed <- read_rds("imported_data/repos_closed.Rds")

repos_closed <- repos_closed %>% 
  select(project=github_org, repo=full_name, description, sponsor=company) %>%
  mutate(commit_access = "closed")

# Commit Log for Repositories (last 6 months)
gitlog_closed_projects <- read_rds("imported_data/commit_log_closed_Jun2018.Rds")

gitlog_closed_projects <- gitlog_closed_projects %>%
  select(project=github_org,
         repo,
         commit.committer.date, 
         # author_name=commit.author.name, author_email=commit.author.email, author_login=author.login,
         committer_name=commit.committer.name, committer_email=commit.committer.email, committer_login=committer.login,
         sha,
         committer_company_name=company) %>%
 mutate(commit_month=floor_date(as.Date(commit.committer.date), unit="months")) %>%
 select(-commit.committer.date) %>%
  filter(commit_month > "2017-12-01")

# Commit Activity for Contributors to the Company Repositories Above
# affliated_commits_closed <- read_rds("imported_data/commit_activity_closed_Jun2018.Rds")
 
```

## Combined Repo List

Make a consolidated list of repos and add information about whether it's a company-owned or public repository.

```{r repos-combined}

repos <- bind_rows(repos_open, repos_closed)

```

## Consolidated Commit Logs

Combine the commit logs and update with metadata about the repos.

```{r gitlog-combined}

# bind rows on commit logs
gitlog <- bind_rows(gitlog_closed_projects, gitlog_open_projects)

```


## Engagement Score

Engagement metrics are defined as follows:

 * Activity Months - Number of months that had at least one commit from an affiliated email address
 * Commit Weight - Rounded natural log of the number of commits from an affiliated email address per month
 
The Engagement Score adjusts the above metrics as follows:

  * Activity Months - Proportion of total months with commits in the range considered for the report. For example, this report analyzes 6 months of data, so the proportion would be the number of months with an affiliated commit out of 6 months.
  * Commit Weight - Takes the mode of the Commit Weight (most frequently occurring Commit Weight over the range analyzed)

Engagement Score = Activity Months * Commit Weight

```{r project-summary}

# consolidate author + committer (ok if counted twice -- this shows an increased interest in the project)

# create a lookup for company type so we don't have to deal with it in the summary below
company_type_lookup <- bind_rows(gitlog %>% select(company_name=author_company_name, company_type=author_company_type), 
                                 gitlog %>% select(company_name=committer_company_name, company_type=committer_company_type))

company_type_lookup <- company_type_lookup %>% unique()

# phase 1 - project, repo, company, type, date, sha
project_companies <-  gitlog %>%
  select(project, repo, author_company_name, committer_company_name, activity_month=commit_month, sha) %>%
  # convert committer_date to month
  unite(company_name, author_company_name, committer_company_name) %>%
  separate_rows(company_name, sep="_", convert=TRUE)

# group by project
project_summary <- project_companies %>%
  group_by(project, repo, company_name, activity_month) %>%
  summarize(commits=n(),
             # natural log provides a simple weight, add 1 so we can multiply later
            commit_weight = round(log(commits)) + 1) %>%
  group_by(project, repo, company_name, commit_weight) %>%
  mutate(commit_weight_mode = n()) # for computing the mode

# project, company, engagement (activity interval)

project_engagement <- project_summary %>%
  group_by(project, company_name) %>%
  summarize(
    activity_months = n_distinct(activity_month),
    commit_weight = commit_weight[which.max(commit_weight_mode)]
  )

activity_months_max <- max(project_engagement$activity_months)

project_engagement <- project_engagement %>% 
  # proportion of months with a commit
  mutate(months_pct = round(activity_months/activity_months_max, 1),
         engagement_score = months_pct * commit_weight) %>%
  # add company type
  inner_join(company_type_lookup)

# add repo data for plotting

project_engagement <- project_engagement %>%
  inner_join(repos %>% select(-repo, -description) %>% unique())

write_csv(project_engagement, "data/project_engagement.csv")
```

# Analysis

## IBM vs Microsoft

Focus specifically on a subset of companies.

```{r engagement-subset}

engagement_ibm <- project_engagement %>%
  filter(! (is.na(company_name) & commit_access == "closed")) %>% # haven't identified authors on closed repos yet
  filter(company_name %in% c("Microsoft", "IBM") | sponsor %in% c("Microsoft", "IBM")) %>%
  mutate(company_type=ifelse(is.na(company_type) & company_name %in% c("Microsoft", "IBM"), "public", company_type),
         sponsor=ifelse(is.na(sponsor), "Community Project", sponsor))

```

### Raw Comparison

```{r fig-raw-engagment-ibm-vs-microsoft, fig.width=10, fig.height=10}

ggplot(engagement_ibm, 
       mapping = aes(x = project, y = engagement_score)) +
  geom_bar(aes(fill = company_name), stat="identity", position="dodge") +
  coord_flip() +
  facet_wrap(~ sponsor, scales="free", ncol=2) +
  labs(x="Project", y="Engagement Level", fill="Contributing Company")
```

```{r fig-raw-engagment-ibm-vs-microsoft-summary, fig.width=10, fig.height=10}

ggplot(engagement_ibm, 
       mapping = aes(x = project, y = engagement_score)) +
  geom_bar(aes(fill = company_name), stat="identity", position="dodge") +
  coord_flip() +
  facet_wrap(~ sponsor, scales="free", ncol=2) +
  labs(x="Project", y="Engagement Level", fill="Contributing Company")

engagement_ibm_summary <- engagement_ibm %>%
  mutate(sponsor=ifelse(sponsor %in% c("Microsoft", "IBM"), sponsor, "External")) %>%
  filter(company_name %in% c("Microsoft", "IBM"))

ggplot(engagement_ibm_summary, 
       mapping = aes(x = project, y = engagement_score)) +
  geom_bar(aes(fill = company_name), stat="identity", position="dodge") +
  coord_flip() +
  facet_wrap(~ sponsor, scales="free_y", ncol=1) +
  labs(x="Project", y="Engagement Level", fill="Contributor")

```




### Engagement Baseline

What does "normal" engagement look like for the companies?

### Adjusted Comparisons

TODO: Include affiliated commits

TODO: Only compare AI/DS "closed" projects.



